Parsing Postfix log files

### 2008-07-22 doc/logparser-architecture.tex
Moved the architecture stuff into it's own section.

\section{Background}

\label{background}

This section provides background information helpful in understanding the
remainder of the document.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
document, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this document.  Other projects which attempt to
parse Postfix log files are summarised (full details are available in
\sectionref{other-parsers}), finishing with a review of previously
published research in this area.

\subsection{Motivation}

\label{motivation}

This document and the program it describes are part of a larger project to
optimise a mail server's Postfix restrictions, generate statistics and
graphs, and provide a platform on which new restrictions can be trialled
and evaluated to determine if they are beneficial in the fight against
spam.  The program parses Postfix log files and populates a database with
the data gleaned from those log files, providing a consistent and simple
view of the log files which future tools can utilise.  The gathered data
can then be used to optimise current anti-spam measures, provide a baseline
to test new anti-spam measures against, or to produce statistics showing
how effective those measures are.

A snippet of \SQL{} provides a short example of the optimisation possible
using data from the database: determining which Postfix restrictions reject
the highest number of mails:

\begin{verbatim}
SELECT name, description, restriction_name, hits_total
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

If the database supports sub-selects percentages can be obtained:
\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
the database containing the extracted data does not support this syntax,
then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
results will be the same, just slightly less visually pleasing.}

\begin{verbatim}
SELECT name, description, restriction_name, hits_total,
        (hits_total * 100.0 /
            (SELECT SUM(hits_total)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

XXX GIVE SAMPLE OUTPUT FOR THE SNIPPET ABOVE\@.

Another example is determining which restrictions are not effective: this
shows which restrictions had fewer than 100 hits on the last log file
parsed, and the percentage of total rejections each restriction represents.

\begin{verbatim}
SELECT name, description, restriction_name, hits,
        (hits * 100.0 /
            (SELECT SUM(hits)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
        AND hits < 100
    ORDER BY hits ASC;
\end{verbatim}

These database queries yield summary statistics about the efficiency of
spam avoidance techniques; statistics that are far less feasible to assess
directly from log files without prior pre-processing into a database in the
fashion proposed, implemented and tested herein.

\subsection{Database as Application Programming Interface}

\label{database as API}

The database populated by this program provides a simple interface to
Postfix log files.  Although the interface is a database schema, it is in
effect quite similar to any other \API{} provided by shared code: it
insulates both user and provider of the \API{} from changes in the
implementation of the \API{}\@.  The algorithm implemented by the parser
can be improved; support can be added for earlier or later releases of
Postfix; bugs can be fixed or limitations removed from the parser; these
changes will not cause the user to be negatively impacted.  Statistics
and/or graphs can be generated from the database; new restrictions can be
tested and the results inspected; trends in the fight against spam can
emerge from historical data saved in the database; the parser remains the
same as the usage adapts.  Using a database simplifies writing programs
which need to interact with the data in several ways:

\begin{enumerate}

    \item It is already possible to access databases from the vast majority
        of programming languages, allowing a developer to access the data
        gathered using their preferred programming language, rather than
        being restricted to the language the parser is written in.  It is
        often possible to write an interface layer allowing code written in
        one language to be used in another language, but this greatly
        increases the effort required to use the parser.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

    \item Databases are easily extensible, e.g.:

            \item Other tables can be added to the database, e.g.\ to cache
                historical, summary or computed data.

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
                for explanations of these terms,
                e.g.~\cite{sql-for-web-nerds}}

            \item A VIEW gives a custom arrangement of data with very
                little effort.

            \item If the database supports it, access can be granted on a
                fine-grained basis, e.g.\ allowing the finance department
                to produce invoices, the helpdesk to run limited queries as
                part of dealing with support calls, and the administrators
                to have full access to the data.

            \item Triggers can be written to perform actions when certain
                events occur.  In pseudo-\SQL{}\@:

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = 'boss@example.com'
        AND postfix_action = 'REJECTED'
    SEND PANIC EMAIL TO 'postmaster@example.com';
\end{verbatim}

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn (a good
        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
        every vendor implements a different dialect of \SQL{}, the basics
        are the same everywhere (analogous to the overall similarities and
        minor differences between Irish English, British English, American
        English and Australian English).  Depending on the database in use
        there may be tools available which reduce or remove the requirement
        to know \SQL{}; for SQLite (the default database used by the
        implementation) there are several available~\cite{sqlite-guis}.

\end{enumerate}

Storing the results in a database will also increase the efficiency of
using those results, as the log files need only be parsed once rather than
each time the data is used; indeed the results may be used by someone with
no access to the original log files.

\subsection{Simple Mail Transfer Protocol}

\label{SMTP background}

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity of the protocol many virus and/or spam sending programs fail
to implement it properly, so requiring strict adherence to the protocol
specification is beneficial in protecting against spam and
viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
adhered to the principle of \textit{Be liberal in what you accept, and
conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
unfortunately that principle was written in a friendlier time.  Given the
deluge of spam that mail servers are subjected to daily, a more appropriate
maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
the strongest restrictions you can; relax the restrictions and adherence
only when legitimate mail is impeded.\/}  it is not as friendly, nor as
catchy, but it more accurately reflects the current situation.} A typical
\SMTP{} conversation resembles the following (the lines starting with a
three digit number are sent by the server, all other lines are sent by the
client):

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: <alice@example.com>
250 2.1.0 Ok
RCPT TO: <bob@example.com>
250 2.1.5 Ok
DATA
354 End data with <CR><LF>.<CR><LF>
Message headers and body sent here.
.
250 2.0.0 Ok: queued as D7AFA38BA
QUIT
221 2.0.0 Bye
\end{verbatim}

An example deviation from the protocol:

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: Alice N. Other alice@example.com
501 5.1.7 Bad sender address syntax
RCPT TO: Bob in Sales/Marketing bob@example.com
503 5.5.1 Error: need MAIL command
DATA
503 5.5.1 Error: need RCPT command
Message headers and body sent here.
.
502 5.5.2 Error: command not recognized
QUIT
221 2.0.0 Bye
\end{verbatim}

This client is so poorly written that not only does it present the sender
and recipient addresses improperly, it ignores the error messages returned
by the server and carries on regardless.  There are many spam and virus
sending programs which are this deficient --- unfortunately others
(particularly newer programs) were written by competent programmers, or
utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
would have done its best to deal with deficient clients, with the intention
of accepting as much mail destined for its users as
possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
command, or accepting sender or recipient addresses which were not enclosed
in \texttt{<>}.

A detailed description of \SMTP{} is beyond the scope of this document:
introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02},
the definitive references are~\cite{RFC821, RFC2821}.

\subsection{Postfix}

\label{postfix background}

Postfix is a \MTA{} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive, extensible, optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of stricture and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained in \sectionref{policy
servers}.} as the ultimate in flexibility.  Administrators can also supply
their own rejection messages to make it clear to senders why exactly their
mail was rejected.  Unfortunately this flexibility has a cost: complexity
in the log files generated.  While it is easy to use standard Unix text
processing utilities to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  For the majority of mails the journey is simple and brief, but
the remaining minority can be quite complex (see \sectionref{additional
complications} for details).

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an
\MTA{}\@: receive mail, send mail, local delivery of mail, etc.\ --- full
details can be found in~\cite{postfix-overview}.  Each log line contains
the name of the Postfix component which produced it, and this information
is used when determining which rules should be used to parse each log line
(see \sectionref{rule characteristics} for details).  This design has
positive security implications also: those components which interact with
other hosts are not privileged,\footnote{Privilege in this case means the
power to perform actions that are limited to the administrator, and not
available to ordinary users.} so bugs in those components will not give an
attacker extra privileges; those components which are privileged do not
interact with other hosts, making it much more difficult for an attacker to
exploit any bugs which may exist in those components.

\subsubsection{Mixing and matching Postfix restrictions}

\label{Mixing and matching Postfix restrictions}

Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
smtpd_per_user_control, policy-servers}; the following is a brief overview
only.

Postfix uses one restriction list (containing zero or more restrictions)
for each stage of the \SMTP{} conversation: client connection, HELO
command, MAIL FROM command, RCPT TO commands, DATA command, and end of
data.  The appropriate restriction list is evaluated for each stage
(evaluation will be explained shortly), though by default the restriction
lists for client connection, HELO and MAIL FROM commands will not be
evaluated until the first RCPT TO command is received, because some clients
do not deal properly with rejections before RCPT TO\@; a benefit of this
delay is that Postfix has more information available when logging
rejections.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible as described
in~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.} The
meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing more
specific cases to be used as exceptions to more general cases.  When the
result is the name of another restriction Postfix will evaluate the new
restriction, allowing restrictions to be chosen based on the client \IP{}
address, HELO hostname, sender address, recipient address,
etc.\footnote{E.g.\ the administrator may require that all clients on the
local network have valid \DNS{} entries, to prevent people sending mail
from unknown machines.}  The administrator can define new restrictions as a
list of existing restrictions, allowing arbitrarily long and complex
sequences of lookups, restrictions and exceptions.  Postfix tries to
protect the administrator in as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
all users --- the administrator must be cognisant of the ramifications of
their configuration changes.  XXX UNIX DOESN'T STOP YOU DOING STUPID THINGS
BECAUSE THAT WOULD ALSO STOP YOU DOING CLEVER THINGS\@.

Postfix uses simple lookup tables as the deciding factor when evaluating
some restrictions, e.g.\ in the example restriction below\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

    \item [check\_client\_access] The name of the restriction to evaluate.

    \item [cidr] The type of the lookup table.

    \item [/etc/postfix/client\_access] The file containing the lookup
        table.

The restriction \texttt{check\_client\_access} checks whether the \IP{}
address of the connected client is found in the file and returned the
associated action if found; the method of searching the file is dependant
on the type of the file (\texttt{cidr} in the example) --- see
\cite{postfix-lookup-tables} for more details.  Other restrictions
determine their result by consulting external sources, e.g.\
\texttt{reject\_rbl\_client dnsbl.example.com} checks whether the \IP{}
address of the client is present in the \DNSBL{}
\texttt{dnsbl.example.com}, rejecting the command if the client is listed.
This description is necessarily brief, for further details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.

\subsubsection{Policy servers}

\label{policy servers}

A \textit{policy server\/}~\cite{policy-servers} is an external program
consulted by Postfix to determine the fate of an \SMTP{} command.  The
policy server is given state information\footnote{Sample state information
is shown in \tableref{Example attributes sent to policy servers}} and
returns a verdict as described in \sectionref{Mixing and matching Postfix
restrictions}.  The policy server can perform more complex checks than
those provided by Postfix: a trivial example is restricting mail from
addresses associated with the payroll system to sending mail on the third
Tuesday after pay day only, to help prevent problems from spam or (worse)
phishing mails with faked sender addresses.\footnote{A phishing mail might
claim that the payroll system had a disastrous disk failure; until the
server is replaced and restored all salary payments will have to be
processed manually, so please reply to this mail with your name, address
and back account details.}

Some widely deployed policy servers:

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
        \SPF{}\label{spf introduction} records specify which mail servers
        are allowed to send mail claiming to be from a particular domain.
        The intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}, e.g.\ forwarding mail from one company or
        university to another when a user changes jobs.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender address, recipient address,
        remote \IP{} address) is unknown; on second and subsequent delivery
        attempts from that triple the mail will be accepted.  The
        assumption is that maintaining a list of failed addresses and
        retrying after a temporary failure is uneconomical for a spam
        sender, but that a legitimate mail server must retry.  Sadly spam
        senders are using increasingly complex and well written programs to
        distribute spam, frequently using an \ISP{} provided \SMTP{} server
        from a compromised machine on the \ISP{}'s network.  Greylisting
        will slowly become less useful, but it does block a large
        percentage of spam mail at the moment; the most effective
        restrictions over the \numberOFlogFILES{} log files used in testing
        the parser are shown in \tableref{Summary of rejections}.
        Greylisting is obviously worth using, at least at the moment,
        particularly when you factor in Greylisting's position as the final
        restriction which a mail must overcome:\footnote{Greylisting is the
        final restriction a mail must overcome in the configuration used on
        the mail server the log files were obtained from; an administrator
        is free to use Greylisting at whichever position in the restriction
        list they feel is most appropriate for their mail system.}
        Greylisting only takes effect for mails which have passed every
        other restriction.

        \begin{table}[ht]
            \caption{Summary of rejections}\label{Summary of rejections}
            \input{build/restriction-table-include.tex}
        \end{table}

    \item Using a scoring system such as
        Policyd-weight~\cite{policyd-weight} where tests accumulate points
        against the sending system --- if the eventual score is higher than
        a threshold the mail is rejected.  Postfix's restrictions are
        binary, and the administrator must manually whitelist clients if
        they are to bypass a restriction; using a threshold which requires
        hitting several restrictions frees the administrator from
        whitelisting clients which fall foul of one restriction only.

    \item Rate limiting on a per-sender, per-client or per-recipient basis
        as performed by Policyd~\cite{policyd}.

Example attributes taken from~\cite{policy-servers}:

\begin{table}[ht]

    \caption{Example attributes sent to policy servers}\label{Example
    attributes sent to policy servers}

    \begin{tabular}[]{ll}

        request                 & smtpd\_access\_policy     \\
        protocol\_state         & RCPT                      \\
        protocol\_name          & SMTP                      \\
        helo\_name              & some.domain.tld           \\
        queue\_id               & 8045F2AB23                \\
        sender                  & foo@bar.tld               \\
        recipient               & bar@foo.tld               \\
        recipient\_count        & 0                         \\
        client\_address         & 1.2.3.4                   \\
        client\_name            & another.domain.tld        \\
        reverse\_client\_name   & another.domain.tld        \\
        instance                & 123.456.7                 \\

    \end{tabular}
\end{table}

\subsection{Assumptions}

The algorithm described and the program implementing it make a small number
of (hopefully safe and reasonable) assumptions:

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log rotation gone awry,
        file system filling up, logging system unable to cope with the
        volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering a deluge of spam or a
        mail loop.

    \item Postfix logs sufficient information to make it possible to
        accurately reconstruct the actions it has taken.  There are a
        number of heuristics used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
        logging after cleanup} for details.

    \item The Postfix queue has not been tampered with, causing unexplained
        appearance or disappearance of mail.  This may happen if the
        administrator deletes mail from the queue without using
        \daemon{postsuper}, or if there is filesystem corruption.

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time: there are 375,750
lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
code.} there are advantages to treating Postfix as a black box while
developing the parser:

    \item The parser is developed using real world log files rather than
        the idealised log files someone would naturally envisage reading
        the source code.

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling independently interferes with logging and other
        processing.

    \item The parser acts as a second source of information, with the
        information gathered from empirical evidence.  An interesting
        project would be to compare the empirical knowledge inherent in the
        parsing algorithm with the documentation and source code of
        Postfix.

@@

\subsection{Conventions used in the document}

The words \textit{connection\/} and \textit{mail\/} are often used
interchangeably in this document; in general the word used was chosen based
on the context it appears in.

\subsection{Summary}

This section has provided background information on several topics relevant
to the remainder of the document.  It started with the motivation behind
the project, continuing with explanations of:

@@

    \item Using a database as an \API{}.

    \item \SMTP{}.

    \item Postfix restrictions and policy servers.

    \item Assumptions made when designing and developing the parser.

    \item A description of the parser's novel design.

    \item Conventions used in this document.

    \item An in-depth comparison of this parser with \LMA{}, the parser
        described in previously published research.

    \item A review of the literature previously published in this area.

@@

>>>>

\section{Parser Architecture}

@@

\subsection{Rule characteristics}

\label{rule characteristics}

Rule have certain characteristics which may help in understanding the
parser:

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules have
        already have been tried and have failed on the current log line.}
        Any given rule will only be used to parse a subset of the log
        lines, and any given log line will only have a subset of the rules
        tried against it.

    \item The first matching rule wins: no further rules are tried against
        that log line, but there is a facility for prioritising the rules
        so that more specific rules can be tried first.

    \item Rules are completely self-contained and can be understood in
        isolation, without reference to any other rules.

    \item Rule processing time is a linear function of the number of rules.
        XXX IMPROVE THIS\@.

@@

\label{comparison against context-free grammars}

In context-free grammar terms the parser rules could be described as:

$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

\subsection{Overlapping rules}

\label{overlapping rules}

XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?

XXX DFA COMPARISON\@.

The parser does not try to detect overlapping rules; that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line may change between log
files, and the first matching rule wins.  Overlapping rules are frequently
a requirement, allowing a more specific rule to match some log lines and a
more general rule to match the majority, e.g.\ separating \SMTP{} delivery
to specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Negative priorities may be useful
for catchall rules.

Detecting overlapping rules is difficult, but the following approaches may
be helpful:

    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
        similar to: \textbf{select regex from rules order by regex;}

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules (described in \sectionref{rule efficiency}).  Parse
        a number of log files using normal sorting, then dump a textual
        representation of the rules, connections and results tables.
        Repeat with shuffled and reversed rules, starting with a fresh
        database.  If there are no overlapping rules the tables from each
        run will be identical; differences indicate overlapping rules.
        Which rules overlap can be determined by examining the differences
        in the tables: each result contains a reference to the rule which
        created it, if the references differ between runs the two rules
        referenced in the differing results overlap.  Unfortunately this
        method cannot prove the absence of overlapping rules; it can detect
        overlapping rules, but only if there are log lines in the input
        files which match more than one rule.

\subsection{Framework}

\label{framework}

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.

This functionality is provided by the framework, which both drives parsing
overall and provides services to the actions it invokes.  The framework is
responsible for loading rules from the database and sanity checking them,
reading log files, matching each rule against the current log line,
invoking the correct action, maintaining state tables, loading and saving
state, displaying a progress indicator, and miscellaneous other tasks.
Actions use services provided by the framework, including storing and
retrieving data in the state tables, extracting and saving data captured by
rules, storage of global data, debugging functions, preparing a mail for
entry in the database, and entering mails in the database.

There is a similarity between this design and the event-driven programming
paradigm commonly used in GUI programs, where one part of the program
responds to events (mouse clicks in a GUI program, log lines being matched
in the parser) and invokes the correct action.

The framework's high level overview could be expressed as (indentation
denoting flow of control):

\begin{verbatim}
for each line in the input files:
    for each rule defined by the user:
        if this rule matches the input line:
            perform the action specified by the rule
            skip the remaining rules
            process the next input line
    warn the user that the input line was not parsed
\end{verbatim}

### 2008-07-22 doc/logparser-background.tex
Moved the architecture stuff into it's own section.

This section provides background information helpful in understanding the
remainder of the document.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
document, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this document.  Other projects which attempt to
parse Postfix log files are summarised (full details are available in
\sectionref{other-parsers}), finishing with a review of previously
published research in this area.

@@

\subsection{Parser design}

\label{parser design}

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix are not fixed; they vary
widely from host to host, depending on the set of restrictions chosen by
the administrator.  With this in mind, one of the parser's design aims was
to make adding new rules as easy as possible, to enable administrators to
properly parse their log files.  To enable this the parser is divided into
three parts:

    \item [rules] Rules match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines with actions, and are described in
        detail in \sectionref{rules}.

    \item [actions] Actions are invoked to deal with a log line once it has
        been identified by the rules: actions modify data structures,
        handle complications, and cause data to be saved to the database.
        The actions perform the work of reconstructing the journey a mail
        takes through Postfix.  Full details of actions can be found in
        \sectionref{actions-in-detail} and \sectionref{adding new actions}
        XXX IMPROVE THIS SENTENCE\@.

    \item [framework] The framework is responsible for loading rules,
        managing data structures, finding the rule which matches each log
        line, invoking the correct action, etc\@.  The framework provides
        the structure which actions and rules plug into.  The framework is
        described in detail in \sectionref{framework}.  XXX EXTEND
        FRAMEWORK SECTION IF NECESSARY\@.

It may help to think of the rules and actions as components which plug into
the framework.

\label{why separate rules, actions and framework?}

XXX MERGE THE NEXT TWO PARAGRAPHS\@.

Decoupling the parsing rules from the associated actions and framework
allows new rules to be written and tested without requiring modifications
to the parser source code (significantly lowering the barrier to entry for
new or casual users who need to parse new log lines), and greatly
simplifies framework, actions and rules.  Decoupling also creates a clear
separation of functionality: rules handle low level details of identifying
log lines and extracting data from a log line; actions handle the higher
level details of following the path a mail takes through Postfix,
assembling the required data before storing it, dealing with complications
arising, etc; the framework provides services to actions and stores data.

Decoupling the actions from the framework simplifies both framework and
actions: the framework provides services to the actions, and does not need
to deal with the complications which arise, or the task of reconstructing a
mail's journey through Postfix; actions benefit from having services
provided by the framework, freeing them to concentrate on the task of
accurately reconstructing each mail's journey through Postfix and dealing
with the complications described in \sectionref{additional complications}.

XXX CLARIFY THE SENTENCE STARTING WITH ``Although this may see'' IN THE
NEXT PARAGRAPH\@.

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change and make the appropriate changes, versus adding a new
rule with the action to invoke, a \regex{} to match the log lines, and the
specification of the data to extract.  Bear in mind that the changes must
be made without adversely affecting existing parsing, particularly as there
may be edge cases which are not immediately obvious.\footnote{See
\sectionref{yet-more-aborted-delivery-attempts} for a complication which
occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
changes to the parser's code also complicates upgrades, as the changes must
be preserved during the upgrade, and may clash with changes made by the
developer.\footnote{See \sectionref{adding new actions} for how to add new
actions}  \parsername{} allows the user to add new rules to the database
without changing the parsing algorithm, unless the new log lines to be
parsed require functionality not already provided by the algorithm.  If the
new log lines do require new functionality, new actions can be added to the
parser without modifying existing actions or other parts of the algorithm;
only in the rare case that the new actions require support from other
sections of the code will more extensive changes be required.

XXX CHANGE THIS\@; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
TOO MUCH\@.

There is some similarity between the parser's design and William Wood's
\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
\ATN{} and the parser is accidental, but it is interesting how two
apparently different approaches share an underlying separation of concerns;
this appears to be a natural division of responsibility and functionality.

% Do Not Reformat!

\begin{tabular}[]{lll}
    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
    Networks            & Algorithm         & Determines the sequence 
                                              of transitions            \\
                        &                   & or actions which 
                                              constitutes a valid       \\
                        &                   & input.                    \\
    Transitions         & Actions           & Save data and impose
                                              conditions the            \\
                        &                   & input must meet to be
                                              considered valid.         \\
    Abbreviations       & Rules             & Responsible for 
                                              classifying input.        \\
\end{tabular}

>>>>

This section provides background information helpful in understanding the
remainder of the document.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
document, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this document.

@@
